---
title: "South Africa Security"
author: "Ivan Musebe"
date: "4/12/2023"
output: html_document
---
# _`South Africa Security Analysis (2015 - 2023)`_

## _`Set Working Directory`_
Set working directory for ease of access of the necessary files.
```{r}
setwd("~/Documents/GitHub/denaco-analytics/R") 
# change to "~denaco-analytics/R" if project downloaded in root folder
```

Install the required libaries for the environment:
```{r}
install.packages(
  c('sf', 'tidyr', 'dplyr', 'tibble', 'leaflet',
  'janitor', 'plotly', 'classInt', 'lubridate', 'IRdisplay', 
  'tidyverse', 'htmlwidgets', 'RColorBrewer', 'leaflet.extras',
  'leaflet.providers'),
  repos='http://cran.us.r-project.org')
```

Load the libraries to make the functions and tools provided by each library available for use in your R code.
Note: Explain the functions of each library
```{r}
# load libraries:
library(sf)
library(tidyr)
library(dplyr)
library(tibble)
library(plotly)
library(leaflet)
library(janitor)
library(classInt)
library(lubridate)
library(IRdisplay)
library(tidyverse)
library(htmlwidgets)
library(RColorBrewer)
library(leaflet.extras)
library(leaflet.providers)
```

## _`Data`_
### Load data:
The data used in the analysis is extracted from the [ACLED](https://acleddata.com/) database. ACLED is a platform that records ongoing live incidents of crime, protests, terrorism and intercommunal violence across the world.
Our data is for South Africa incidents dated January 01 2015 - March 31 2023
```{r}
df <- read.csv(
  "South_Africa_Security_Incidents 2015-2023.csv",
  header = TRUE, na.strings = ""
)
```

### About the Data
#### Structure of the data
To know the structure of the data we will use the function `str()` to know the variables in our object and the data types of each variable.
```{r}
str(df)
```
Our data contains the security incidents information;
  a. Where,
  b. When,
  c. What &
  d. How an incident occured
The data also shows the actors in an incidents and the fatalities in each incident,if any.
This is the data we are going to analyze.

## Cleaning Data
### Missing values
The [ACLED Codebook](https://acleddata.com/acleddatanew/wp-content/uploads/2021/11/ACLED_Codebook_v1_January-2021.pdf) guides us on the key variables.
As noted on the structure of the data "actor 2", "assoc_actor_2" have have a large number of missing values as well as "civil targeting".
We have 6610 records. Searching through for their actial figures might prove time consuming and hence we will procede and drop them asa method of dealing with missing values.
```{r}
# Make a copy of the data
df_copy <- data.frame(df)

# subset droping the columns
df_copy <- subset(df_copy, select = -c(11, 12, 15))
```

### Feature engineering
#### Dates
The dates column is written in two formarts
  1. %d-%m-%y &
  2. %d %M %Y for the months of April
Using the code below, format the date to a single format:
```{r}
# Function that changes multiple date formats to a single format
multidate <- function(data, formats){
    a<-list()
    for(i in 1:length(formats)){
        a[[i]]<- as.Date(data,format=formats[i])
        a[[1]][!is.na(a[[i]])]<-a[[i]][!is.na(a[[i]])]
        }
    a[[1]]
    }

# Format the date variable (event_date)
df_copy$event_date <- multidate(
  df_copy$event_date,
  # input the date formats in the data
  c("%d-%b-%y","%d %B %Y")
)
```
Using the dates, now in the right format we can have timeseries charts of security trends.

From the date variable we can produce the month,week of the year (woy) and day of the week.
```{r cars}
df_copy <- df_copy %>%
            add_column(
                month = format(df_copy$event_date, "%b"),
                week = paste(lubridate::isoweek(df_copy$event_date), sep = ""),
                day = weekdays(df_copy$event_date, abbreviate = TRUE),
                .after = "year"
            )
```

#### Geocodinates
Use the Google geocoding API to map the longitudes and latitudes of each location.
```{r}
# _Input geocode API key
register_google(key = "xxxxxxxxxx") #(For subscription)

Co
Address <- df_copy[, c("location", "country")] %>% unite("Address" , c("location", "country"), sep = ",", na.rm = TRUE)

for(i in 1:nrow(df_copy)){
  loc_coord <- data.frame(
    tryCatch(
      geocode(Address$Address[i], output = "latlona",source = "google"),
      warning = function(w) data.frame(lon = NA, lat = NA)
    )
  )
  df_copy$latitude[i] <- as.numeric(loc_coord[2])
  df_copy$longitude[i] <- as.numeric(loc_coord[1])
      
}
```

View clean data
```{r}
head(df_copy)
```
 
 ### Save data
```{r}
write.csv(df_copy, file = "clean_SA_incidents 2015-2023", row.names = FALSE)
```

## _`Statistics & Visualizations`_
In this section we will derive descriptive statistics and visualize them.
One can analyse the data using the following real time questions

### 1. What is the trend of security incidents in South Africa?
```{r}
df_copy %>%
    group_by(month = lubridate::floor_date(event_date, 'month'))  %>%
    summarise(total_count =n(), .groups = 'drop') %>%
  plot_ly(x = ~month, y = ~total_count, type = "scatter",
          text = ~total_count, textposition = "upper right",
          line = list(color = c("red")),
          marker = list(color = c("#2c1429")),
          width = 650, height = 500) %>%
    layout(plot_bgcolor="#f6f6f6",
           paper_bgcolor="#f6f6f6",
          #  font = list(family = "calibri", size = 14),
           xaxis = list(title = "",
                        zeroline = FALSE),
           yaxis = list(title = "Count",
                        zeroline = FALSE))
```

### 2. How many incidents occured each year?
```{r}
df_copy %>%
    group_by(year)  %>%
    summarise(total_count =n(), .groups = 'drop') %>%
  plot_ly(x = ~as.factor(year), y = ~total_count, type = "bar",
          text = ~total_count, textposition = "outside",
          marker = list(color = c("red")),
          width = 650, height = 500) %>%
    layout(plot_bgcolor="#f6f6f6",
           paper_bgcolor="#f6f6f6",
          #  font = list(family = "calibri", size = 14),
           xaxis = list(title = "Year",
                        zeroline = FALSE),
           yaxis = list(title = "Count",
                        zeroline = FALSE))
```